---
title: "EDA-challenge"
author: "Erika Carlson"
date: 2024-02-07
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
editor_options: 
  chunk_output_type: console
---

# Explore and Wrangle Data
Load the “data-wrangling.csv” dataset from [**this URL**](https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/data-wrangling.csv) as a tabular data structure named **d** and look at the variables it contains


```{r}
library(tidyverse)
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/data-wrangling.csv"
d <- read_csv(f, col_names = TRUE)
names(d)
```

Create new variables
- **BSD** (body size dimorphism), the ratio of mean male to female body mass
- **sex_ratio**, the ratio of the number of adult females to adult males in a typical group
- **DI** (for “defensibility index”), the ratio of day range length to the diameter of the home range
```{r}
d <- d %>% 
  mutate(BSD = Body_mass_male_mean/Body_mass_female_mean, 
         sex_ratio = AdultFemale/AdultMales,
         DI = DayLength_km/(sqrt(HomeRange_km2/pi)*2)) # sqrt(Area/pi)*2 = d

```


Plot the relationship between **day range length** and **time spent moving**, for these primate species overall and by family. 
- Do species that spend more time moving travel farther overall? *A: Yes, based on regression line*
- How about within any particular primate family? *A: Yes for Atelidae, Hylobatidae, Cebidae*
- Should you transform either of these variables? *A: log transformed day range length*

```{r}
library(ggplot2)
library(cowplot) # to show graphs side by side
plot1 <- ggplot(data = d, 
            aes(x = log(DayLength_km), 
                y = Move)) +  # build a plot object
  xlab("log(Day range length)") + ylab("Time spent moving") + # modify the axis labels
  geom_point(na.rm = TRUE) + # make a scatterplot
  geom_smooth(method = "lm", na.rm = TRUE) + # add a regression line
  ylim(0, 40) + # set y-axis range
  theme(legend.title = element_blank()) # modify the legend
plot1 # plot the object

plot2 <- ggplot(data = d, 
            aes(x = log(DayLength_km), 
                y = Move,
                color = factor(Family))) +  # build a plot object and color points by Family
  xlab("log(Day range length)") + ylab("Time spent moving") + # modify the axis labels
  geom_point(na.rm = TRUE) + # make a scatterplot
  geom_smooth(method = "lm", na.rm = TRUE) + # add a regression line
  ylim(0, 40) + # set y-axis range
  theme(legend.title = element_blank()) # modify the legend
plot2 # plot the object

plot_grid(plot1, plot2, rel_widths = c(1, 1.5), labels = "AUTO")
```


Plot the relationship between **day range length** and **mean group size**, overall and by family. 
- Do species that live in larger groups travel farther overall? *A: Yes, based on regression line*
- How about within any particular primate family? *A: Yes for Cercopithecidae, Hominidae, Tarsiidae, Cebidae*
- Should you transform either of these variables? *A: log transformed day range length*

```{r}
plot1 <- ggplot(data = d, 
            aes(x = log(DayLength_km), 
                y = MeanGroupSize)) +  # build a plot object
  xlab("log(Day range length)") + ylab("Group Size") + # modify the axis labels
  geom_point(na.rm = TRUE) + # make a scatterplot
  geom_smooth(method = "lm", na.rm = TRUE) + # add a regression line
  ylim(0, 100) + # set y-axis range
  theme(legend.title = element_blank()) # modify the legend
plot1 # plot the object

plot2 <- ggplot(data = d, 
            aes(x = log(DayLength_km), 
                y = MeanGroupSize,
                color = factor(Family))) +  # build a plot object and color points by Family
  xlab("log(Day range length)") + ylab("Group Size") + # modify the axis labels
  geom_point(na.rm = TRUE) + # make a scatterplot
  geom_smooth(method = "lm", na.rm = TRUE) + # add a regression line
  ylim(0, 100) + # set y-axis range
  theme(legend.title = element_blank()) # modify the legend
plot2 # plot the object

plot_grid(plot1, plot2, rel_widths = c(1, 1.5), labels = "AUTO")
```


Plot the relationship between **body size dimorphism** and **canine size dimorphism** overall and by family. 

Do taxa with greater size dimorphism also show greater canine dimorphism? *A: Yes, overall and in Carcopithecidae, Hominidae, Cebidae, Tarsiidae, and Lorisidae families*

```{r}
plot1 <- ggplot(data = d, 
            aes(x = BSD, 
                y = Canine_Dimorphism)) +  # build a plot object
  xlab("Body size dimorphism") + ylab("Canine size dimorphism") + # modify the axis labels
  geom_point(na.rm = TRUE) + # make a scatterplot
  geom_smooth(method = "lm", na.rm = TRUE) + # add a regression line
  # ylim(0, 100) + # set y-axis range
  theme(legend.title = element_blank()) # modify the legend
plot1 # plot the object

plot2 <- ggplot(data = d, 
            aes(x = BSD, 
                y = Canine_Dimorphism,
                color = factor(Family))) +  # build a plot object and color points by Family
  xlab("Body size dimorphism") + ylab("Canine size dimorphism") + # modify the axis labels
  geom_point(na.rm = TRUE) + # make a scatterplot
  geom_smooth(method = "lm", na.rm = TRUE) + # add a regression line
  # ylim(0, 100) + # set y-axis range
  theme(legend.title = element_blank()) # modify the legend
plot2 # plot the object

plot_grid(plot1, plot2, rel_widths = c(1, 1.5), labels = "AUTO")
```


Create a new variable named diet_strategy that is “frugivore” if fruits make up >50% of the diet, “folivore” if leaves make up >50% of the diet, and “omnnivore” if neither of these is true. Then, do boxplots of group size for species with different dietary strategies. 

Do frugivores live in larger groups than folivores? *A: No, overlap in IQR (frugivore has more outliers)*

```{r}
d <- d %>% 
  mutate(diet_strategy = case_when(Fruit > 50 ~ "frugivore",
                                   Leaves > 50 ~ "folivore",
                                   Fruit < 50 & Leaves < 50 ~ "omnivore")) # specify to avoid NAs
```

```{r}
p <- ggplot(data = d, 
            aes(x = factor(0), y = MeanGroupSize)) + 
  geom_boxplot(na.rm = TRUE, outlier.shape = NA) + 
  theme(axis.title.x = element_blank(), 
        axis.text.x = element_blank(), 
        axis.ticks.x = element_blank()) + 
  geom_dotplot(binaxis = "y", stackdir = "center", 
               stackratio = 0.2, alpha = 0.3, dotsize = 0.5, color = NA, 
               fill = "blue", na.rm = TRUE) +
  facet_grid(. ~ diet_strategy) + geom_rug(sides = "l")
p
```


In one line of code, using {dplyr} verbs and the forward pipe (`%>%` or `|>`) operator, do the following:
- Add a variable, **Binomial** to the data frame **d**, which is a concatenation of the **Genus** and **Species**…
- Trim the data frame to only include the variables **Binomial**, **Family**, **Brain_size_species_mean**, and **Body_mass_male_mean**…
- Group these variables by **Family**…
- Calculate the average value for **Brain_size_species_mean** and **Body_mass_male_mean** per **Family** (remember, you may need to specify na.rm = TRUE)…
- And arrange by increasing **average brain size**

```{r}
d <- d %>% 
  mutate(Binomial = str_c(Genus, " ", Species)) %>% 
  select(Binomial, Family, Brain_Size_Species_Mean, Body_mass_male_mean) %>% 
  group_by(Family) %>% 
  summarize(avg_brain_size = mean(Brain_Size_Species_Mean),
            avg_male_body_mass = mean(Body_mass_male_mean)) %>% 
  arrange(avg_brain_size)
```

